---
title: 'Homework II: SVD analysis & Life Tables'
subtitle: 'Deadline: 2020-05-20'
author: "Duc Nguyen DUONG & Mohamed-Amine BOUSAHIH"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_caption: yes
    keep_md: yes
    number_sections: yes
    toc: yes
  html_notebook:
    code_folding: none
    number_sections: yes
    toc: yes
params:
    dirpath: '../LIFE_TABLES/lt_'
    country_code: 'fr_t'
    country: 'France'
    timecourse: 1945:2015
---


---

# Objectives

This notebook aims at

  - working with **tables** (`data.frames`, `tibbles`, `data.tables`, ...) using `dplyr` or any other query language (as provided for example by `data.table`)
  - visualizing demographic data as provided by [Human Mortality Database organization](https://www.mortality.org).
  - using PCA and other matrix oriented methods (CCA) to explore a multivariate datasets (lifetables may be considered as multivariate datasets)

# Life tables data (ETL)

Life data tables are downloaded from [https://www.mortality.org](https://www.mortality.org).

See also [https://www.lifetable.de](https://www.lifetable.de).

If you install and load package [https://cran.r-project.org/web/packages/demography/index.html](https://cran.r-project.org/web/packages/demography/index.html), you will also find life data tables.

We investigate life tables describing countries from Western Europe (France, Great Britain --actually England and Wales--, Italy, the Netherlands, Spain, and Sweden) and the United States.

We load the one-year lifetables for female, male and whole population for the different countries.

```{r, echo=FALSE, eval=FALSE}
# for debugging
# params should be initialized from YAML header
params<- list(
    timecourse= 1945:2015,
    dirpath= '../LIFE_TABLES/lt_',
    country_code= 'fr_t',
    country= 'France')
```

```{r, echo=FALSE}
timecourse <- eval(rlang::parse_expr(params$timecourse))
```

```{r tidyverse, message=FALSE, warning=FALSE, echo=FALSE}
pacman::p_load(tidyverse)
pacman::p_load(plotly)
pacman::p_load(foreach)
pacman::p_load(iterators)
pacman::p_load(DT)
pacman::p_load(ade4)
pacman::p_load(FactoMineR)
pacman::p_load(factoextra)
pacman::p_load(FactoInvestigate)
pacman::p_load(ggfortify)

old_theme <-theme_set(theme_dark(base_size=9,
                                 base_family = "Helvetica"))

knitr::opts_chunk$set(eval=FALSE,
  echo=FALSE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  autodep = TRUE,
  tidy = FALSE)

```


```{r, echo=FALSE}
country_code <- list(fr_t='FRATNP',
                     fr_c='FRACNP',
                     be='BEL',
                     gb_t='GBRTENW',
                     gb_c='GBRCENW',
                     nl='NLD',
                     it='ITA',
                     swe='SWE',
                     sp='ESP',
                     us='USA')

countries <- c('fr_t',  'gb_t',  'nl', 'it', 'sp', 'swe', 'us')

country_names <- list(fr_t='France',     # total population
                     fr_c='France',      # civilian population
                     be='Belgium',
                     gb_t='England & Wales',    # total population
                     gb_c='England & Wales',    # civilian population
                     nl='Netherlands',
                     it='Italy',
                     swe='Sweden',
                     sp='Spain',
                     us='USA')

gender_names <- list('b'='Both',
                     'f'='Female',
                     'm'='Male')
```


Check on [http://www.mortality.org](http://www.mortality.org) the meaning of the different columns:

Document [Tables de mortalité françaises pour les XIXe et XXe siècles et projections pour le XXIe siècle](https://www.lifetable.de/data/FRA/FRA000018061997CY1.pdf) contains
detailed information on the construction of Life Tables for France.
Two kinds of Life Tables can be distinguished: *Table du moment* which
contain for each calendar year, the mortality risks at different ages for that very year; and *Tables de génération* which contain for a given birthyear, the   mortality risks
at which an individual born during that year has been exposed. The life tables investigated in this homework are *Table du moment*. According to the document by Vallin and Meslé, building the life tables  required decisions and doctoring.

See (among other things)

- p. 19 Abrupt changes in mortality quotients at some ages for a given calendar year
- Estimating mortality quotients at great age.

Have a look at [Lexis diagram](https://en.wikipedia.org/wiki/Lexis_diagram).

Some columns need retyping:

- `Year`: should be integer
- `Age`:  needs some cleaning, after cleaning it should be typed as integer
- `Lx`: should be integer
- `Tx`: should be integer
- Other columns should be considered as floating point numbers (`numeric`)

---

- [ ] After ETL processing, you obtain a universal table encompassing
all data available  in the files located in `LIFE_TABLES` directory.
Henceforth, the universal table is named `life_table`, its schema should be
the following.

| Column Name | Column Type |
|:------------|:------------|
|  Year       | integer     |
|  Age        | integer     |
|  mx         | double      |
|  qx         | double      |
|  ax         | double      |
|  lx         | integer     |
|  dx         | integer     |
|  Lx         | integer     |
|  Tx         | integer     |
|  ex         | double      |
|  Country    | factor      |
|  Gender     | factor      |

Coercion introduces a subtantial number of NA  warnings.
Preliminary inspection of the data suggests that coercion problems
orginate from column `Age`: `110+` cannot be coerced to an integer value.
Discard corresponding rows using `tidyr::drop_na(Age)`.

```{r}
fusion_data = function(data_f,data_m,country)
{
  data_f["Gender"] = "F"
  data_m["Gender"] = "M"
  data = rbind(data_f,data_m)
  data["Country"] = country
  return (data)
}

gender_code = list('b'='both',
                  'f'='female',
                  'm'='male')

import_data = function(country,gender)
{
  file = paste("../LIFE_TABLES/lt_",gender_code[gender],"/",gender,"ltper_1x1/",country,".",gender,"ltper_1x1.txt",sep="")
  data = readr::read_delim(file, delim = ' ', skip = 2)
  return(data)
}

retype = function(data)
{
  names(data) <- stringr::str_replace_all(names(data), " ", "")
  data %>%
  dplyr::mutate_at(vars(Year,Age,lx,dx,Lx,Tx),.funs = list(~as.integer(.))) %>%
  dplyr::mutate_at(vars(Country,Gender),.funs = list(~forcats::as_factor(.))) %>%
  dplyr::mutate_at(vars(mx,qx,ax,ex),.funs = list(~as.double(.)))-> data
  return(data)
}

universal_table = function()
{
  index = (2:length(countries))
  data_f = import_data(country_code[countries[1]],'f')
  data_m = import_data(country_code[countries[1]],'m')
  data = fusion_data(data_f,data_m,country_names[countries[1]])
  for (c in countries[index])
  {
    data_f = import_data(country_code[c],'f')
    data_m = import_data(country_code[c],'m')
    temp = fusion_data(data_f,data_m,country_names[c])
    data = rbind(data,temp)
  }
  return(data)
}

df = universal_table()
df = retype(df)
df%>%tidyr::drop_na(Age)->df

glimpse(df)
```

# Western countries in 1948

- [ ] Plot the central death rates of all Countries at all ages for year 1948.
```{r}
p<- ggplot(mapping=aes(x=Age, y=mx, colour = Country))+
  geom_point(size = 0.5)+
  xlab("Age") +
  ylab("Mortality quotients") +
  ggtitle("Central death rates of all Countries at all ages for year 1948")+
  scale_x_continuous(breaks = seq(0,110,by=10))+
  scale_y_log10()+
  facet_wrap(~ Gender)+
  theme_bw()

(p%+%
  dplyr::filter(df,Year == 1948))%>%ggplotly()


```

- [ ] Comment.
- [ ] Plot ratios between central death rates in European countries and
central death rates in the USA in 1948.```
```{r}
ratio <- function(data, reference_country="USA", reference_year=1948){
  data%>%
    dplyr::filter(Year == reference_year)->data
  
  dplyr::filter(data, Country != reference_country)%>%
  dplyr::inner_join(y=data[data$Country==reference_country, c("Age", "mx","Gender")], by=c("Age","Gender"))
}

q <- ggplot(mapping = aes(x=Age, y=mx.x/mx.y, colour = Country)) +
  geom_smooth(method="loess",
              se=FALSE,
              span=0.2,
              size=.5) +
  ggtitle("Ratios of central death rates between European countries and USA in 1948")+
  ylab("Ratio of mortality rates in 1948, reference Country USA") +
  scale_y_continuous(breaks = seq(0,7,by=1))+
  scale_x_continuous(breaks = seq(0,110,by=10))+
  facet_wrap(~ Gender)+
  theme_bw()

(q %+%
  ratio(df))%>%ggplotly()

```
- [ ] Comment

# Death rates evolution since WW II

- [ ] Plot mortality quotients (column `mx`) for
both genders as a function of `Age` for  years `1946, 1956, ...` up to `2016` .
Use aesthetics to distinguish years.
```{r}
after_ww_II = seq(1946,2016,by=10)
p<-
  ggplot(mapping = aes(x=Age, y=mx, colour = forcats::as_factor(Year)))+
  geom_smooth(method="loess",
              se=FALSE,
              span=0.2,
              size=.5) +
  scale_y_log10()+
  labs(colour="Year")+
  xlab("Age") +
  ylab("Mortality quotients") +
  theme_bw()
(p%+%
  dplyr::filter(df, Year %in% after_ww_II)+
  ggtitle("Mortality quotients as function of age for years 1946, 1956, ... 2016"))%>%ggplotly()
```

- [ ] Facet by `Gender`  and `Country`
```{r}
(p%+%
  dplyr::filter(df, Year %in% after_ww_II) +
  facet_wrap(vars(Gender,Country)))%>%ggplotly()
```

- [ ] Write a function `ratio_mortality_rates` with signature
`function(df, reference_year=1946, target_years=seq(1946, 2016, 10))`
that takes as input:

  - a dataframe with the same schema as `life_table`,
  - a reference year `ref_year` and
  - a sequence of years `target_years`

and that returns a dataframe with schema:


| Column Name | Column Type |
|:------------|:------------|
|  Year       | integer     |
|  Age        | integer     |
|  mx         | double      |
|  mx.ref_year| double      |
|  Country    | factor      |
|  Gender     | factor      |

where `(Country, Year, Age, Gender)` serves as a _primary key_,
`mx` denotes the central death rate at `Age` for `Year` and `Gender` in `Country`
whereas `mx_ref_year` denotes central death rate at `Age` for argument `reference_year`
in `Country` for `Gender`.


```{r}
ratio_mortality_rates <- function(df,
                                  reference_year=1946,
                                  target_years=seq(1946, 2016, 10))
{
  dplyr::filter(df[c("Age","mx","Country","Gender","Year")], Year %in% target_years) %>%
  dplyr::inner_join(y=df[df$Year==reference_year, c("Age", "mx", "Gender","Country")], by=c("Age","Gender","Country"))->data
  return (data);
}

ratio = ratio_mortality_rates(df)

```


- [ ] Draw plots displaying the ratio  $m_{x,t}/m_{x, 1946}$ for ages $x \in 1, \ldots, 90$
and year $t$ for $t \in 1946, \ldots, 2016$ where $m_{x,t}$ is the central death rate
at age $x$ during year  $t$.

  1. Handle both genders and countries `Spain`, `Italy`, `France`, `England & Wales`, `USA`, `Sweden`, `Netherlands`.
  1. One properly facetted plot is enough.
```{r}
p<- ggplot(mapping = aes(x=Age, y=mx.x/mx.y, colour = Country)) +
  geom_smooth(method="loess",
              se= FALSE,
              size =.5,
              span= .2)+
  scale_x_continuous(breaks = seq(0,110,by=10))+
  ylab("Ratio of mortality rates, reference Year 1946") +
  ggtitle("Variation of mortality quotient with respect to Y=1946")+
  facet_wrap(~ Gender)+
  theme_bw()

(p%+%
  dplyr::filter(ratio, Age <= 90))%>%ggplotly()
```

- [ ] Comment


# Trends


- [ ] Plot mortality quotients at ages $0, 1, 5$  as a function of time. Facet by Gender and Country

```{r}
ages <- c(0, 1, 5)
p <-ggplot(mapping=aes(x=Year, y=mx, linetype=forcats::as_factor(Age))) +
  geom_smooth(method="loess",
              se=FALSE,
              span=0.2,
              size=.5) +
  scale_color_brewer() +
  labs(linetype="Age") +
  facet_wrap(vars(Gender,Country))+
  ylab("Mortality quotients")+
  theme_bw()
(p%+%
  filter(df, Age %in% ages)+
  ggtitle("Mortality quotients at ages 0, 1, 5 as a function of time"))
```

- [ ] Comment

- [ ] Plot mortality quotients at ages $15, 20, 40, 60$  as a function of time. Facet by `Gender` and `Country`

```{r}
ages <- c(15, 20, 40, 60)
p%+%
  filter(df, Age %in% ages)+
  ggtitle("Mortality quotients at ages 15, 20, 40, 60 as a function of time")
```

- [ ] Comment

# Rearrangement

- [ ] From dataframe `life_table`, compute another dataframe  called `life_table_pivot`
with primary key `Country`, `Gender` and `Year`, with a column for each `Age` from `0` up to `110`.
For each age column, the entry should be the central death rate
at the age defined by column, for `Country`, `Gender`  and `Year` identifying the row.

You may use functions `pivot_wider`, `pivot_longer` from `tidyr::` package.

The resulting schema should look like:

| Column Name | Type    |
|:------------|:--------|
| Country     | factor  |
| Gender      | factor  |
| Year        | integer |
| `0`         | double  |
| `1`         | double  |
| `2`         | double  |
| `3`         | double  |
| $\vdots$    | $\vdots$|
```{r}
df_pivot <- df %>%
  select(Year, Age, Gender,Country, mx) %>%
  tidyr::pivot_wider(names_from = "Age", values_from = "mx")
```
- [ ] Using `life_table_pivot` compute life expectancy at birth for each Country, Gender and Year
```{r}
ex <- function(mx){
  sum(cumprod(1 - mx))
}
df_pivot %>%
  dplyr::select(Year, Gender, Country) -> df_ex
df_ex["life_expectancy"]= apply(df_pivot[,-c(1,2,3)], MARGIN=1, FUN = ex)

#compare to real ex
df %>%
  filter(Age==0) %>%
  select(Year, Gender,Country,ex)
df_ex
```

# Life expectancy

- [ ] Write a function that takes as input a vector of mortality quotients, as well as an age, and returns the residual life expectancy corresponding to the vector and the given age.
```{r}
rex <- function(mx, age){
  sum(cumprod(1 - mx[(age+1):length(mx)]))
}

```
- [ ] Write a function that takes as input
a dataframe with the same schema as  `life_table` and returns
a data frame with columns `Country`, `Gender`, `Year`, `Age` defining a primary key
and a column `res_lex` containing _residual life expectancy_ corresponding to the pimary key.

In order to compute residual life expectancies, you may consider using `window` functions
oer apropriately defined windows. Package `dplyr` does not offer a rich API for window functions.
Package `dbplyr` does.

```{r}
life_expectancy_table = function(lt)
{
  lt %>%
    select(Country, Gender,Year, Age, mx) %>%
    group_by(Year, Country, Gender) %>%
    summarise(mx_list = list(mx))%>%
    inner_join(lt[c("Age","Country","Gender","Year")],c("Year", "Gender","Country"))%>%
    group_by(Year, Country, Gender, Age) %>%
    summarise(ex = rex(unlist(mx_list),Age))%>%
    arrange(Country, Gender) -> lt
  return (lt)
}

df_ex_for_all_age = life_expectancy_table(df)
```
- [ ] Plot residual life expectancy as a function of `Year` at ages $60$ and $65$,
facet by `Gender` and `Country`.

```{r retirement, echo=FALSE, eval=FALSE}
ages = c(60,65)
p<-ggplot(mapping=aes(x=Year, y=ex, linetype=forcats::as_factor(Age))) +
  geom_smooth(method="loess",
              se=FALSE,
              span=0.2,
              size=.5) +
  scale_color_brewer() +
  labs(linetype="Age") +
  ylab("Life expectancy")+
  ggtitle("Life expectancy as a function of Year at ages 60 and 65")+
  facet_wrap(vars(Gender,Country))+
  theme_bw()
p%+%
  filter(df_ex_for_all_age, Age %in% ages)
```



# PCA  and SVD over log-mortality tables

- [ ] Pick a Country, a Gender,  a range of years `1948:2010`.
Extract the corresponding  lines from `life_table_pivot`. Take logarithms of
central death rates and perform principal component analysis.
```{r}
library(tidyverse)
library(broom)
years = c(1948:2010)
extract_data_pca = function(lt,ref_country,ref_gender,years)
{
  lt%>%
    filter(Country == ref_country, Gender == ref_gender, Year %in% years) -> lt
  data.frame(lt[,1],apply(lt[,-c(1,2,3)],MARGIN = 2,FUN = log)) -> lt
  return(lt)
}
lt_pca = extract_data_pca(df_pivot,"Spain","M",years)
pca = prcomp(lt_pca,scale = TRUE)
```
```{r}
# Screeplot - V2

screeplot(pca, type = "l", npcs = 15, main = "Screeplot")
abline(h = 1, col="red", lty=5)
legend("topright", legend=c("Eigenvalue = 1"),
       col=c("red"), lty=5, cex=0.6)
```
```{r}
# Cumulative variance plot 

cumpro <- cumsum(pca$sdev^2 / sum(pca$sdev^2))
plot(cumpro[0:15], xlab = "Principal Components", ylab = "Amount of explained variance", main = "Cumulative variance plot")
abline(v = 2, col="blue", lty=5)
abline(h = 0.915, col="blue", lty=5)
legend("topleft", legend=c("Cut-off at PC2"),
       col=c("blue"), lty=5, cex=0.6)


```
```{r}
#Corcicle

COR = cor(lt_pca,pca$x[,1:2])
s.corcircle(COR)
```
```{r}
#Biplot

augment(pca, data = lt_pca)%>%
  ggplot(aes(.fittedPC1, .fittedPC2)) +
  geom_point() +
  geom_text(aes(label = Year))+
  ggtitle("Biplot")+
  theme_bw()
```

- [ ] comment the screeplot
- [ ] comment the correlation circle
- [ ] comment the biplot

# Canonical Correlation Analysis

- [ ] Build a function that takes as input
  + a dataframe like `life_table_pivot`,
  + a couple  of countries, say `Spain` and `Sweeden`,
  + a vector of `Year`, say `1948:1998`
  + a `Gender` say `Female`
  returns a matrix called $Z$ with rows corresponding to `Year` and columns
  corresponding to couples `(Country, Age)`.

- [ ] Perform a Canonical Correspondance Analysis of $Z$
```{r}
library(vegan)
extract_data_cca = function(lt,ref_country,ref_gender,years)
{
  lt%>%
    filter(Country == ref_country, Gender == ref_gender, Year %in% years) -> mat
  mat = as.data.frame(mat)
  mat %>% remove_rownames %>% column_to_rownames(var="Year") -> mat
  mat = mat[,-c(1,2,3)]
  return(mat)
}
spain_f_cca = extract_data_cca(df_pivot,"Spain", "F", c(1948:1998))
usa_f_cca = extract_data_cca(df_pivot,"USA", "F", c(1948:1998))

ccamodel = cca(spain_f_cca~.,usa_f_cca)
finalmodel<- ordistep(ccamodel, scope=formula(ccamodel))
finalmodel
anova.cca(finalmodel)
```
```{r}
plot(finalmodel,type = "n",main="CCA")
text(finalmodel,dis = "cn")
text(finalmodel,pch = 21, col = "red", bg = "yellow", cex = 0.5)
text(finalmodel,"species",col = "blue",cex = 0.5)
```
- [ ] Comment ...
- [ ] Comment ...

# Lee-Carter model for US mortality

During the last century, in the USA and in western Europe,
central death rates at all ages have exhibited a general decreasing trend.
This decreasing trend has not always been homogeneous across ages.

The Lee-Carter model has been designed to model and forecast the
evolution of the log-central death rates for the United States during the XXth century.

Let $A_{x,t}$ denote the log central death rate at age $x$ during year $t\in T$
for a given population (defined by Gender and Country).

The Lee-Carter model assumes that observed loagrithmic central death rates
are sampled according to the following model
\[
A_{x,t} \sim_{\text{independent}} a_x + b_x \kappa_t + \epsilon_{x,t}
\]
where $(a_x)_x, (b_x)_x$ and $(\kappa_t)_t$ are unknown vectors that satisfy
\[
a_x = \frac{1}{|T|}\sum_{t \in T} A_{x,t}\qquad \sum_{t\in T} \kappa_t = 0 \qquad \sum_{x} b_x^2 =1
\]
and $\epsilon_{x,t}$ are i.i.d Gaussian random variables.

## US data

- [ ] Fit a Lee-Carter model on the American data (for Male and Female data) training on years `1933` up to `1995`.
- [ ] Compare the fit provided by the Lee-Carter model with the fit provided by a rank $2$
truncated SVD
- [ ] Compare vectors avec $(a_x)_x, (b_x)_x$ and $(\kappa_t)_t$ with appropriate singular vectors.
- [ ] Use the Lee-Carter model to predict the central death rates for years $2000$ up to $2015$
- [ ] Plot predictions and observations for years $2000, 2005, 2010, 2015
```{r}
df_lc = df %>%
  select(Year,Age,Gender,Country,mx)

k_forcaster = function(k,year,last_training_year,drift,var){
  res = tail(k,1)
  for (i in (1:(year-last_training_year)))
  {
    res = res + drift + rnorm(1, 0, var) 
  }
  return (as.numeric(res))
}

lee_carter = function(df_lt,training_country,training_years,fitting_country,fitting_years, gender)
{
  lt <- df_lc %>%
  filter(Country == training_country, Year %in% training_years, Gender == gender) %>%
  select(Year, Age, Gender,Country, mx) %>%
  tidyr::pivot_wider(names_from = "Age", values_from = "mx")
  
  data.frame(lt[,c(1,2,3)],apply(lt[,-c(1,2,3)],MARGIN = 2,FUN = log)) -> lt
  M <- lt %>%
  select(-c(Year,Gender,Country))
  # The mean log-rate for each age group.
  a <- colMeans(M)
  # Subtract the average age pattern a from all years
  for (j in 1:110)
  {
    M[,j] <- M[,j] - a[j]
  }
  # Compute SVD, M = U D V' where U and V' are orthogonal matrices and D is diagonal matrix of singular values. This vector models how to different age groups react to mortality change.
  d <- svd(M, 1, 1)
  # Lee and Carter normalize the first row of V so it sums to one and call it b.
  b <- d$v/sum(d$v)
  # Lee and Carter take the first column of U, multiply by D_1,1 and multiply by the sum of the first row of V' and call that k. This vector captures overall mortality change over time.
  
  k <- d$u * sum(d$v) * d$d[1]
  
  #k_data <- cbind(lt$Year,k) 
  #k_data = as.data.frame(k)
  #reg = lm(V2~V1,data = k)
  #coef = reg$coefficients
  #eq = paste("k = ",round(coef[2],1), "* t + ",round(coef[1],1))
  #plot(k_data,ylab = "K", xlab="Year")
  #abline(reg)
  drift = (tail(k,1) - k[1])/length(k)
  var_k = c()
  for (i in (1:(length(k)-1)))
  {
    var_k = append(var_k,var(c(k[i],k[i+1])))
  }
  var_k = mean(var_k)
  mx_lc = c()
  for (i in fitting_years)
  {
    mx_lc = append(mx_lc,exp(a + b*k_forcaster(k,i,tail(training_years,1),drift,0.5)))
  }
  result <- filter(df_lc, Year %in% fitting_years, Country == fitting_country, Gender == gender) %>%
  mutate(Year = factor(Year), mx_lc)
  return(result)
}



```
```{r}
training_years = c(1933:1995)
fitting_years = c(2000:2015)
training_country = "USA"
fitting_country = "USA"
gender = "F"

usa_lc_00_15 = lee_carter(df_lc,training_country,training_years,fitting_country,fitting_years,gender)
# Plotting parameters and fits.
ptheme = theme(
  panel.background = element_rect(fill = "white"),
  panel.border = element_rect(fill = NA, colour = "black",size = 0.25),
  panel.grid.major = element_line(colour = "grey83",size = 0.25),
  panel.grid.minor = element_line(colour = "grey83",size = 0.25),
  legend.justification = c(1,0),
  legend.position = c(1,0)
)

cols = c("c1" = "red", "c2" = "blue")
p<- ggplot(mapping=aes(x = Age)) + 
  geom_point(aes(y = mx, color = "c1"),size = 0.3, alpha= .7) + 
  geom_point(aes(y = mx_lc, color = "c2"),size = 0.3, alpha= .7) + 
  scale_color_manual(name = "Legend" ,breaks= c("c1","c2"),values = cols, labels = c("Observation","Prediction"))+
  ylab("Mortality quotient")+
  facet_wrap(~Year)+
  scale_y_log10()+
  ptheme+
  theme_bw()
```
```{r}
years = seq(2000,2015,by = 5)
(p%+%
  filter(usa_lc_00_15,Year %in% years)+
  ggtitle(paste("Predictions of mortality of quotient using Lee Carter model for",fitting_country,"trained by", training_country,"data")))
```

## Application of Lee-Carter model to a European Country

- [ ] Fit a Lee-Carter model to a European country
```{r}
training_country = "USA"
training_years = c(1933:1995)
fitting_country = "Spain"
fitting_years = c(2000:2015)
years = seq(2000,2015,by = 5)
gender = "M"
spain_lc_00_15 = lee_carter(df_lc,training_country,training_years,fitting_country,fitting_years,gender)
```
```{r}
(p%+%
  filter(spain_lc_00_15,Year %in% years)+
  ggtitle(paste("Predictions of mortality of quotient using Lee Carter model for",fitting_country,"trained by", training_country,"data")))
```
```{r}
training_country = "Spain"
training_years = c(1933:1995)
fitting_country = "Spain"
fitting_years = c(2000:2015)
years = seq(2000,2015,by = 5)
gender = "M"
spain_lc_00_15 = lee_carter(df_lc,training_country,training_years,fitting_country,fitting_years,gender)
```
```{r}
(p%+%
  filter(spain_lc_00_15,Year %in% years)+
  ggtitle(paste("Predictions of mortality of quotient using Lee Carter model for",fitting_country,"trained by", training_country,"data")))
```
- [ ] Comment
- [ ] Compare with rank-2 truncated SVD
- [ ] Use the Lee-Carter model to predict the central death rates for years $2000$ up to $2015$
      Plot predictions and observations for years $2000, 2005, 2010, 2015$

## Predictions of life expectancies at different ages

- [ ] Use Lee-Carter approximation to approximate residual life expectations
```{r}
life_expectancy_lc = function(lc)
{
  lc %>%
    group_by(Year, Country, Gender) %>%
    summarise(mx_list = list(mx_lc))%>%
    inner_join(lc[c("Age","Country","Gender","Year")],c("Year", "Gender","Country"))%>%
    group_by(Year, Country, Gender, Age) %>%
    summarise(ex_lc = rex(unlist(mx_list),Age))%>%
    arrange(Country, Gender) -> lc
  
  return (lc)
}


```
- [ ] Compare with observed residual life expectations
```{r}
life_expectancy_real = function(df,years,country,gender)
{
  df %>%
    filter(Year %in% years, Country == country, Gender == gender)%>%
    select(ex) -> result
  return(as.matrix(result))

}
ptheme = theme(
  panel.background = element_rect(fill = "white"),
  panel.border = element_rect(fill = NA, colour = "black",size = 0.25),
  panel.grid.major = element_line(colour = "grey83",size = 0.25),
  panel.grid.minor = element_line(colour = "grey83",size = 0.25),
)
p<- ggplot(mapping=aes(x = Age)) + 
  geom_point(aes(y = ex, color = "c1"),size = 0.3, alpha= .7) + 
  geom_point(aes(y = ex_lc, color = "c2"),size = 0.3, alpha= .7) + 
  scale_color_manual(name = "Legend",breaks= c("c1","c2"),values = cols, labels = c("Observation","Prediction"))+
  ylab("Life expectancy")+
  facet_wrap(~Year)+
  scale_y_log10()+
  ptheme+
  theme_bw()
```
```{r}
years = c(2000:2015)
country = "USA"
gender = "F"

lc_usa_ex = life_expectancy_lc(usa_lc_00_15) 
lc_usa_ex$ex = life_expectancy_real(df,years,country,gender)

years = seq(2000,2015,by = 5)
(p%+%
  filter(lc_usa_ex,Year %in% years)+
  ggtitle(paste("Predictions of life expectancy using Lee Carter model for",country)))
```
```{r}
years = c(2000:2015)
country = "Spain"
gender = "M"

lc_spain_ex = life_expectancy_lc(spain_lc_00_15) 
lc_spain_ex$ex = life_expectancy_real(df,years,country,gender)

years = seq(2000,2015,by = 5)
(p%+%
  filter(lc_spain_ex,Year %in% years)+
  ggtitle(paste("Predictions of life expectancy using Lee Carter model for",country)))

```

# References

__Life tables and demography__

- [Human Mortality Database](https://www.mortality.org)
- [Tables de mortalité françaises, Jacques Vallin et France Meslé](https://www.lifetable.de/data/FRA/FRA000018061997CY1.pdf)
- [Modeling and Forecasting U.S. Mortality, R.D.Lee and L.R. Carter, JASA 1992]
- [Les dimensions de la mortalité, S. Ledermann, Jean Breas, Population, 1959]

__Graphics and reporting__

- [Interactive web-based data visualization with R, plotly, and shiny](https://plotly-r.com/index.html)
- [R for Data Science](https://r4ds.had.co.nz)
- [Layered graphics](http://vita.had.co.nz/papers/layered-grammar.pdf)
- [Plotly](http://plotly.com/)

__Tidyverse__

- [tidyselect](https://tidyselect.r-lib.org/articles/tidyselect.html)
- [dbplyr](https://cran.r-project.org/web/packages/dbplyr/vignettes/dbplyr.html)
- [data.table](https://github.com/Rdatatable/data.table)
- [DT](https://rstudio.github.io/DT/)

__PCA, SVD, CCA__

- [FactoMineR](http://factominer.free.fr/index_fr.html)
- [ade4](http://pbil.univ-lyon1.fr/ade4/accueil.php)
- [FactoInvestigate](http://factominer.free.fr/reporting/index_fr.html)
- [PCA and Tidyverse](https://cmdlinetips.com/2019/05/how-to-do-pca-in-tidyverse-framework/)
- [tidyprcomp](https://broom.tidyverse.org/reference/tidy.prcomp.html)
